---
title: "Assignment 2: Coding Basics"
author: "Aditi Jackson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq(1,30,3) # generating sequence of numbers from 1 to 30, increasing by 3's
one_to_thirty_by_three <- seq(1,30,3) # assigning name to sequence
one_to_thirty_by_three # running variable that contains sequence created above

#2. 
mean(one_to_thirty_by_three) # calculating mean of sequence from step 1
median(one_to_thirty_by_three) # calculating median of sequence from step 1

#3. 
mean(one_to_thirty_by_three) > median(one_to_thirty_by_three) # testing if mean > median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

#5 & $6.
Name_vector <- c("Anna", "Ben", "Cameron", "Denis") # Vector with student names (string)
Test_score_vector <- c(99,49,97,48) # Vector with test scores (integer)
Passed_failed_vector <- c(TRUE,FALSE,TRUE,FALSE) # Vector noting passage or failure (Boolean)

#7.
df_student_scores <- as.data.frame(cbind(Name_vector,Test_score_vector,Passed_failed_vector)) #combining vectors into a data frame and assigning it a name
df_student_scores 

#8.
df_student_scores_COPY <- df_student_scores # making copy of original data set in order to apply new column names
df_student_scores_COPY # copy of data
names(df_student_scores_COPY) <- c("Student Name","Test Score","Passed Test? ") # using names() to change column headers
df_student_scores_COPY

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame contains vectors with different modes, which makes it different from a matrix. A data frame is more general than a matrix. While a data frame can comprise vectors with different modes (numbers and characters), all vectors in a matrix must contain the same mode (only numbers or only characters).

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

#10.
x <- Test_score_vector # renaming test score vector as a single letter variable for easy manipulation
passing_grade_function <- function(x){
  if(x>=50){
    print("TRUE - Passing grade")
  }
  else{
    print("FALSE - Failing grade")
  }
}

passing_grade_function(99)


grade_function <- function(x){
  ifelse(x>=50,"TRUE - Passing grade","FALSE - Failing grade")
}

grade_function(60)
grade_function(40)


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: They both worked. They are essentially do the same thing, but ifelse allows you to consolidate the logic
