---
title: "Assignment 3: Data Exploration"
author: "Aditi Jackson"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
getwd() # confirming that working directory is my ENV872 Setup folder
```

```{r installing and loading packages }
# installing packages
install.packages("dplyr") 
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("lubridate")

# loading packages
library("dplyr")
library("ggplot2")
library("tidyverse")
library("lubridate")
```

```{r import data }
# Importing data sets
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = TRUE)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Given that neonicotinoids are widely uses in agriculture, they likely have a huge impact on insect populations and broader ecosystem health. Given that insects play a critical role in most ecosytems, a decline or disrupiton in insect populations could have environemtnal knock-on effects. Studying this data set could help us identify and better understand environmental externalities of using neonictoinoids.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We may be interested in studying litter and woody debris for a variety of different reasons, including gaining insight into forest biodiversity or conducting fire risk assessments. Overall this data can help inform better forest management practices.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and debris sampling is executed at sites that conitan woody vegetation greater than 2 meters tall; samples are collected using a collected of ground and elevated traps
 2. Sampling only occurs in tower plots which are randomly selected
 3. Ground traps are sampled once a year while elevated traps are samples one time every two weeks

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Neonics data dimensions}
dim(Neonics) # dim()  gives use the number of rows and columns in data set
# Neonics has 4623 rows and 30 columns
View(Neonics) # View() allows me to view the entire set of data
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Summary of Neonics Effects}
summary(Neonics$Effect) #  summarizing the effects column of Neonics; the "$" allows us to pull a specific column
```

> Answer: The most commonly studdied effects include Population and Mortality. Population is likely of interest to researchers so that they can understand the effects of neonicotinoids on current and histroical insect populations and identify trends. Similarly, researchers are likely interested in studying the effects of neonicotinoids on mortality to identify the effectivites of insecticdes.

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Summary of Neonics Common Names}
# Summary_CommonName <- summary(Neonics$Species.Common.Name)
# sort(Summary_CommonName)
summary(sort(Neonics$Species.Common.Name)) #embedded sort within summary

```

> Answer: The most commonly studied species are the Honey Bee, the Parasitic Wasp, the Buff Tailed Bumblebee, the Carniolan Honey Bee, the Bumble Bee, and the Italian Honeybee. All six belong to the Hymenoptera, which is a large order of insect species. They are all pollinators and play a crucial role in polinating agricultural crops.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Class of Concentrations}
class(Neonics$Conc.1..Author.) # class of Conc.1..Author. is a factor
```

> Answer: The column above is a factor. This could be because the column was a string in the CSV file, and when we imported the data we told R to "import strings as factors." Therefore our code changed the column to a factor. It's likely not numeric because it is nominal data??

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r graph of studies by year 1}
ggplot(Neonics) + # setting the canvas
  geom_freqpoly(aes(x = Publication.Year), bins=40) # adding paint

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r graph of studies by year 2}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year,color=Test.Location), bins=40) 

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r graph of endpoint counts}

ggplot(Neonics, aes(x = Endpoint)) + # using ggplot to specify data set and column
  geom_bar() + # specifying that I want a bar graph
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # rotating x-axis labels for easier viewing

```

> Answer: The most common end points are NOEL and LOEL. Per the appendix, NOEL is defined as "No-observable-effect-level" and represents :the highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test". LOEL is defined as the "lowest-observable-effect-level" and represents the "lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls."

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r class of collectDate}

class(Litter$collectDate) # class of collectDate is a factor

Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") # converting collectDate column from factor to date using as.Date function

class(Litter$collectDate) # column is now classified as date

unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Niwot Ridge plots}
View(Litter) # viewing data set
unique(Litter$plotID) # unique values in the plotID column. There were 12 unique plots sampled
summary(Litter$plotID) # summary of values in the plotID column 

```

> Answer: Unique tells you the number of unique objects in the column while summary tells you how many of each unique object is in the column.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Graph of FunctionalGroup counts}

ggplot(Litter, aes(x = functionalGroup)) + # using ggplot to specify data set and column
  geom_bar() # specifying that I want a bar graph

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Boxplot of functionalGroup and dryMass}
ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot()
```

```{r Violin graph of funtionalGroup and dryMass}
ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin()
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is a more effective option because the values in the data are very close together, so we are better able to see the distribution of data within a small range. Since there is not a huge spread, the violin plot is ineffective at telling us anything about the data.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles tend to have the highest biomass as we can see from the boxplot.

