---
title: "Assignment 10: Data Scraping"
author: "Aditi Jackson"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
# installing and loading 'tidyverse' and 'rvest' packages
#install.packages(tidyverse)
#install.packages(rvest)
#install.packages(lubridate)
library(tidyverse)
library(rvest)
library(lubridate)

# checking working directory is set to "ENV872 Setup (local)
getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
# setting URL to be scraped
URL_scrape <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3
# scarping water system name
waterSystemName <- URL_scrape %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
waterSystemName

# scarping PWSID
PWSID <- URL_scrape %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
PWSID

# scraping Ownership
ownership <- URL_scrape %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)")%>%
  html_text()
ownership

# scraping Maximum Daily Use (MGD)
maxDailyUse <- URL_scrape %>%
  html_nodes("th~ td+ td") %>%
  html_text()
maxDailyUse

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 

#creating a dataframe with Month, Year, and Max Daily Value
scraped_durham <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                         "Year" = rep(2022,12),
                         "Maximum.Daily.Use"=as.numeric(maxDailyUse))  %>%
  # creating pipe, adding remaining scrapped data as columns and creating a date column
  mutate(Water.System.Name = !!waterSystemName,
         PWSID = !!PWSID,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))

#5
ggplot(scraped_durham)+
  geom_line(aes(x=Date,y=Maximum.Daily.Use))+
  labs(title=paste0("Maximum Daily Water Usage (Durham)"),
     y="Maximum Daily Water Usage",
      x="Date")+
  scale_x_date(date_labels = "%b", date_breaks = "1 month")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 

#Constructing the scraping web address
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
the_PWSID <- '03-32-010'
the_year <- 2022
the_scrape_url <- paste0(the_base_url, 'pwsid=', the_PWSID, '&year=', the_year)
print(the_scrape_url)

# get website's contents
the_website <- read_html(the_scrape_url)

#Creating scraping function
scraping.function <- function(the_year,the_PWSID){

  # Getting website contents
  the_website <- read_html(the_base_url, 'pwsid=', the_PWSID, '&year=', the_year)

  # setting element address variables
  waterSystemName_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  PWSID_tag <-"td tr:nth-child(1) td:nth-child(5)"
  ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  maxDailyUse_tag <-".fancy-table:nth-child(31) th+ td"
  
  # scrapping data items
  waterSystemName <- the_website %>% html_nodes(waterSystemName_tag) %>% html_text()
  PSWID <- the_website %>% html_nodes(PWSID_tag) %>% html_text()
  ownership <- the_website %>% html_nodes(ownership_tag) %>% html_text()
  maxDailyUse <- the_website %>% html_nodes(maxDailyUse_tag) %>% html_text()
  
  # converting to data frame
  df_maxDailyUse <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                         "Year" = rep(the_year,12),
                         "Maximum Daily Use" = as.numeric(maxDailyUse))%>%
      mutate(Water.System.Name = !!waterSystemName,
         PWSID = !!PWSID,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))
    
  return(df_maxDailyUse)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
maxDailyUse2015_df <- scraping.function(2015,'03-32-010')
view(maxDailyUse2015_df)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
